The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-08 rational.tex

To get a feeling for the main idea, consider the following algebraic
manipulation relating a natural number $a$ to itself:
\[\begin{array}{rclll}
&& a & \mbox{initial~$a$} & (0) \\
&=& a + 0 & \mbox{$0$~is~identity~for~$+$} & (1) \\
&=& a + (-a + a) & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (2) \\
&=& (a + (-a)) + a & \mbox{$+$~is~associative} & (3) \\
&=& 0 + a & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (4) \\
&=& a & \mbox{$0$~is~identity~for~$+$} & (5)
\end{array}\]
Although seemingly pointless, this algebraic proof corresponds, in our model,
to an isomorphism of type {{a <-> a}} with a non-trivial and interesting
computational interpretation. The witness for this isomorphism is a
computation that takes a value of type~$a$ (say \$20.00) and eventually
produces the same \$20.00 value as its output. As the semantics of
Sec.~\ref{sec:rat} formalizes, the intermediate steps of the computation
proceed as follows:
\begin{itemize}
\item We start at line (0) with \$20.00; 
\item We proceed to line (1) with the same \$20.00 but tagged as being in the
  left summand of the type $a+0$; we indicate this value as {{left 20}}
\item We continue to line (2) with the same value {{left 20}};
\item At line (3), as a result of re-association the tag on the \$20.00
  changes to indicate that it is in the left-left summand, i.e., the value is
  now {{left (left 20)}};
\item At line (4), we find ourselves needing to produce a value of type 0
  which is impossible; this signals the beginning of a reverse execution
  which sends us back to line (3) with a value {{left (right 20)}};
\item Execution continues in reverse to line (2) with the value 
  {{right (left 20)}};
\item At line (1) we find ourselves again facing an empty type so we reverse
  execution again; we go to line (2) with a value {{right (right 20)}};
\item We proceed to line (3) and (4) with the value {{right 20}};
\item We finally reach line 5 with the value {{20}}.
\end{itemize}

>>>>

To get a feeling for the main idea, consider the following algebraic
manipulation relating a natural number $a$ to itself:
\[\begin{array}{rclll}
&& a & \mbox{initial~$a$} & (0) \\
&=& a + 0 & \mbox{$0$~is~identity~for~$+$} & (1) \\
&=& a + (-a + a) & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (2) \\
&=& (a + (-a)) + a & \mbox{$+$~is~associative} & (3) \\
&=& 0 + a & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (4) \\
&=& a & \mbox{$0$~is~identity~for~$+$} & (5)
\end{array}\]
\[\begin{array}{rlllllllll}
 \!&\!\!  a \!&\! \!&\!       \!&\! \!&\!    \!&\! \mbox{initial~$a$} \!&\! (0) \\
=\!&\!\!  a \!&\!+\!&\!       \!&\!0\!&\!    \!&\! \mbox{$0$~is~identity~for~$+$} \!&\! (1) \\
=\!&\!\!  a \!&\!+\!&\! (-a   \!&\!+\!&\! a) \!&\! \mbox{$-a$~is~the~additive~inverse~of~$a$} \!&\! (2) \\
=\!&\!\! (a \!&\!+\!&\! (-a)) \!&\!+\!&\! a  \!&\! \mbox{$+$~is~associative} \!&\! (3) \\
=\!&\!\!    \!&\!0\!&\!       \!&\!+\!&\! a  \!&\! \mbox{$-a$~is~the~additive~inverse~of~$a$} \!&\! (4) \\
=\!&\!\!    \!&\! \!&\!       \!&\! \!&\! a  \!&\! \mbox{$0$~is~identity~for~$+$} \!&\! (5)
\end{array}\]

Although seemingly pointless, this algebraic proof corresponds, in our model,
to an isomorphism of type {{a <-> a}} with a non-trivial and interesting
computational interpretation. The witness for this isomorphism is a
computation that takes a value of type~$a$ (say \$20.00) and eventually
produces the same \$20.00 value as its output. As the semantics of
Sec.~\ref{sec:rat} formalizes, the intermediate steps of the computation
proceed as follows:
\begin{itemize}
\item We start at line (0) with \$20.00; 
\item We proceed to line (1) with the same \$20.00 but tagged as being in the
  left summand of the type $a+0$; we indicate this value as {{left 20}}
\item We continue to line (2) with the same value {{left 20}};
\item At line (3), as a result of re-association the tag on the \$20.00
  changes to indicate that it is in the left-left summand, i.e., the value is
  now {{left (left 20)}};
\item At line (4), we find ourselves needing to produce a value of type 0
  which is impossible; this signals the beginning of a reverse execution
  which sends us back to line (3) with a value {{left (right 20)}};
\item Execution continues in reverse to line (2) with the value 
  {{right (left 20)}};
\item At line (1) we find ourselves again facing an empty type so we reverse
  execution again; we go to line (2) with a value {{right (right 20)}};
\item We proceed to line (3) and (4) with the value {{right 20}};
\item We finally reach line 5 with the value {{20}}.
\end{itemize}
